Live Functional Programming with Typed Holes

### 2018-04-16 calculus.tex
revisions of sec3 through 3.4 and additional proofs and definitions in the appendix

\parahead{Overview} The syntax of the core calculus given in Fig.~\ref{fig:hazelnut-live-syntax} consists of types and expressions with holes. 
We distinguish between {external} expressions, $e$, and {internal} expressions, $d$. 
External expressions correspond to programs as entered by the programmer 
(see Sec.~\ref{sec:intro} for discussion of implicit, manual, semi-automated and fully automated hole entry methods). 
Each well-typed external expression (as specified in Sec.~\ref{sec:external-statics} below) expands to a well-typed internal expression (see Sec.~\ref{sec:expansion}) before it is evaluated (see Sec.~\ref{sec:evaluation}). 
We take this approach, notably also taken in the ``redefinition'' of Standard ML by \citet{Harper00atype-theoretic}, because (1) the external language supports type inference and explicit type ascriptions, $\hexp : \htau$, but it is formally simpler to eliminate ascriptions and specify a type assignment system when defining the dynamic semantics; and 
(2) we need additional syntactic machinery during evaluation for tracking hole closures and dynamic type casts. 
This machinery is inserted by the expansion step, rather than entered explicitly by the programmer. 
In this regard, the internal language is analogous to the cast calculus in the gradually typed lambda calculus \cite{DBLP:conf/snapl/SiekVCB15,Siek06a}, though as we will see the \HazelnutLive internal language goes beyond the cast calculus in several respects. We have mechanized these formal developments using the Agda proof assistant \cite{norell:thesis,norell2009dependently} 
(see Sec.~\ref{sec:agda-mechanization}). Rule names, written \rulename{SVar}, correspond to the names used in the mechanization. We have also implemented \HazelnutLive in a manner that maintains a novel end-to-end well-definedness invariant (see Sec.~\ref{sec:implementation}).

@@

Before describing the rules in detail, let us state two other guiding theorems. 
The following theorem establishes that an expansion exists for every well-typed external expression. 
The mechanization also establishes when an expansion exists, it is unique (not shown).
 \begin{thm}[Expandability] \label{thm:expandability}~
  \begin{enumerate}[nolistsep]
    \item
      If $\hsyn{\hGamma}{\hexp}{\htau}$
      then $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$
      for some $\dexp$ and $\Delta$.
    \item
      If $\hana{\hGamma}{\hexp}{\htau}$
      then $\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$
      for some $\dexp$ and $\htau'$ and $\Delta$.
  \end{enumerate}
\end{thm}
% \noindent
% The following theorem establishes that when an expansion exists, it is unique.
% \begin{thm}[Expansion Unicity] \label{thm:expansion-unicity}~
%   % \begin{enumerate}[nolistsep]
%   %   \item
%       If $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$
%       and $\expandSyn{\hGamma}{\hexp}{\htau'}{\dexp'}{\Delta'}$
%       then $\htau=\htau'$ and $\dexp=\dexp'$ and $\Delta=\Delta'$.
%   %   \item
%   %     If $\expandAna{\hGamma}{\hexp}{\htau_1}{\dexp}{\htau_2}{\Delta}$
%   %     and $\expandAna{\hGamma}{\hexp}{\htau_1}{\dexp'}{\htau_2'}{\Delta'}$
%   %     then $\dexp=\dexp'$ and $\htau_2=\htau_2'$ and $\Delta=\Delta'$.
%   % \end{enumerate}
% \end{thm}
% \noindent
The following theorem establishes that expansion generalizes external typing.
\begin{thm}[Expansion Generality] \label{thm:expansion-generality}~
  \begin{enumerate}[nolistsep]
    \item
      If $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$
      then $\hsyn{\hGamma}{\hexp}{\htau}$.
    \item
      If $\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$
      then $\hana{\hGamma}{\hexp}{\htau}$.
  \end{enumerate}
\end{thm}

@@

To recap, the result of expansion is a well-typed internal expression with appropriately initialized hole closures and casts. 
We can now specify, in Figures~\ref{fig:isGround}-\ref{fig:step}, the dynamic semantics of \HazelnutLive as a ``small-step'' transition system equipped with a meaningful notion of type safety even for incomplete programs, i.e. expressions typed under a non-empty hole context, $\Delta$. 
We establish that evaluation does not stop immediately when it encounters a hole, nor when a cast fails, by precisely characterizing when evaluation \emph{does} stop. 
We also establish that an essentially standard notion of type safety holds when running complete programs.

@@

The dynamic semantics consists of ,
%
where cast-related machinery closely follows the cast calculus from
the ``refined'' account of the gradually typed lambda calculus
by \citet{DBLP:conf/snapl/SiekVCB15}, which is known to be
theoretically well-behaved.
%
In particular, \Figref{fig:isGround} defines the judgement
$\isGround{\htau}$, which distinguishes the base type~$b$ and the
least specific arrow type~$\tarr{\tehole}{\tehole}$, as \emph{ground
types}; this judgement helps simplify the treatment of function casts.
%
\Figref{fig:isFinal} defines the judgement $\isFinal{d}$, which
distinguishes the final forms of the transition system.
%
There are two classes of final forms: (possibly-)boxed values and
indeterminate forms.%
%
\footnote{
        In most accounts of the cast calculus, values and ground types
        are distinguished with separate grammars together with an
        implicit identification convention.
        %
        Our judgmental formulation is more faithful to the mechanization and
        cleaner for our purposes, because we are distinguishing several
        classes of final forms.
}

The judgement $\isBoxedValue{d}$ defines (possibly-)boxed values as
either ordinary values~(Rule~\rulename{BVVal}), or one of two cast forms: casts
between function types (except identity casts) and casts from a ground
type to the hole type. In each case, the cast operates on a boxed
value.
%
%, which correspond to the values from the cast calculus and include
%the classic values from the lambda calculus, distingui%shed by
%$\isValue{d}$,
%
The judgement $\isIndet{d}$ defines \emph{indeterminate} forms, so
named because they arise from the presence of expression holes and
failed casts in a program, and so, conceptually, their ultimate value
awaits programmer action (see Sec.~\ref{sec:resumption}).
%
The first two rules specify that \emph{empty} hole closures are always
indeterminate, and that \emph{non}-empty hole closures are indeterminate when
they consist of a \emph{final} inner expression.
%
Below, we discuss the the remaining indeterminate forms simultaneously
with the corresponding transition rules, also discussed
below.\todo{relationship to weak head normal forms}

The transition rules are defined in Fig.~\ref{fig:instruction-transitions}-\ref{fig:step}.
%
Top-level transitions are \emph{steps}~$\stepsToD{}{d}{d'}$, governed by Rule~\rulename{Step} in \Figref{fig:step}, which
%
(1) decomposes $d$ into an evaluation context, $\evalctx$, and a selected sub-term, $d_0$;
%
(2) takes an \emph{instruction transition}, $\reducesE{}{d_0}{d_0'}$, as specified in \Figref{fig:instruction-transitions};
%
and (3) places $d_0'$ back in the selected position, indicated in
the evaluation context by the \emph{mark}~$\evalhole$, to obtain $d'$.%
\footnote{
        We say ``mark'' here to avoid confusion with the (orthogonal) ``holes'' of \HazelnutLive.
        %
        %the form $\evalhole$ in the grammar of
        %evaluation contexts is referred to as the \emph{hole}, but
        %this hole is a technical device entirely orthogonal to the
        %holes of this paper, so we use the term ``mark'' instead.
        }
%

%% \matt{Next paragraph can be dropped entirely for space}
%% %
%% This approach was originally developed in the reduction semantics of \citet{DBLP:journals/tcs/FelleisenH92} and is the predominant style of operational semantics in the literature on gradual typing. 
%% Because we distinguish final forms judgementally, rather than syntactically, we use a judgemental formulation of this approach called a \emph{contextual dynamics} by \citet{pfpl}. 
%% It would be straightforward to derive an equivalent structural operational semantics \cite{DBLP:journals/jlp/Plotkin04a} by using search rules instead of evaluation contexts (\citet{pfpl} relates the two approaches).
%% \todo{Put the search rules in the appendix?}{}

@@

\subsubsection{Application and Substitution} 
%
Rule \rulename{ITBeta} in Fig.~\ref{fig:instruction-transitions} specifies the
standard beta reduction transition.
%
The bracketed premises of the form $\maybePremise{\isFinal{\dexp}}$ in
Fig.~\ref{fig:instruction-transitions}-\ref{fig:step} may be \emph{included}
to specify an eager, left-to-right evaluation strategy, or \emph{excluded} to
leave the evaluation order unspecified.
%
In our metatheory, we use the latter version of the rules, both for
the sake of generality, and to support fill-and-resume~(\Secref{sec:resumption}).

A substitution~$[d/x]d'$ operates in the standard capture-avoiding manner~\cite{pfpl}. 
%
The only cases of special interest arise when substitution reaches a hole closure:
\[
\begin{array}{rcl}
  [d/x]\dehole{u}{\sigma}{} & = & \dehole{u}{[d/x]\sigma}{} \\%
  \substitute{d}{x}{\dhole{d'}{u}{\sigma}{}} & = & \dhole{[d/x]d'}{u}{[d/x]\sigma}{}
\end{array}
\]
In both cases, we write~$[d/x]\sigma$ to perform substitution on each expression in the hole environment~$\sigma$. 
%
For example, $\stepsToD{}
    {\hap{(\halam{x}{b}{\halam{y}{b}{\dehole{u}{[x/x, y/y]}{}}})}{c}}
    {\halam{y}{b}{\dehole{u}{[c/x, y/y]}{}}}$. 
%
Reduction may duplicate hole closures, just as with any other substituted term;
%
consequently, the environments of different closures for the same hole name may differ, 
e.g., when an reduction applies a function with a hole closure body multiple times.%
\todo{show example here or refer back to sec 2?}{} 
%
Hole closures may also appear within the environments of other hole
closures, giving rise to the closure paths described in
Sec.~\ref{sec:paths}\todo{where?}{}.

@@

The ITBeta rule is not the only rule we need to handle function
application, because lambdas merely one (final) form of arrow type.
%
Two other situations may also arise.

@@

Second, the expression in function position may be indeterminate,
where arrow cast conversion is not applicable,
e.g. $\hap{(\dehole{u}{\sigma}{})}{c}$.
%
In this case, the application is indeterminate (Rule~\rulename{IAp}
in \Figref{fig:isFinal}), and the application's reduction reduces no
further.

@@

The rules take care to maintain the following property, which establishes that expressions that step are not final, and that final expressions do not step.%
\todo{prove this as stated from Ian's disjointness proofs}{}%
\todo{do we need a typing premise?}{}
\begin{thm}[Finality] There does not exist $d$ such that both $\isFinal{d}$ and $\stepsToD{}{d}{d'}$ for some $d'$.
\end{thm}

\subsubsection{Casts}
Rule \rulename{ITCastId} drops identity casts. The remaining instruction
transition rules assign meaning to non-trivial casts.
%
As discussed in Sec.~\ref{sec:cast-insertion}, the structure of a term
cast to hole type is statically obscure,
%
so we must wait for \emph{use} of the term at some other type, via a
cast away from hole type, to detect dynamic type errors.
%
Rules {ITCastSucceed} and {ITCastFail} handle this situation when the
two types involved are ground types (Fig.~\ref{fig:isGround}).
%
If the two ground types are equal, then the cast succeeds and the cast
may be dropped.
%
If they are not equal, then the cast fails and the failed cast form,
$\dcastfail{\dexp}{\htau_1}{\htau_2}$, arises.
%
Rule \rulename{TAFailedCast} specifies that a failed cast is well-typed exactly
when $d$ has ground type $\tau_1$ and $\tau_2$ is a ground type
disequal to $\tau_1$.
%
Rule \rulename{IFailedCast} specifies that a failed cast operates as an
indeterminate form once $d$ is final.
%
This rule allows evaluation to proceed after cast failure, much as
with hole closures.

Rules {TAFailedCast} and {IFailedCast} each operate at ground type. 
%
The two remaining instruction transition rules, Rule~\rulename{ITGround}
and~{ITExpand}, insert intermediate casts from non-ground type to a
consistent ground type, and \emph{vice versa}.
%
These rules serve as technical devices, permitting us to restrict our
interest exclusively to ground types.
%
Here, the only non-ground types are the arrow types, so the grounding
judgement~$\groundmatch{\tau_1}{\htau_2}$ (\Figref{fig:groundmatch}),
produces the ground arrow type~$\tarr{\tehole}{\tehole}$.
%
More generally, this judgement is governed by the following invariant:
\begin{lem}[Grounding] 
  If $\groundmatch{\htau}{\htau'}$
  then $\isGround{\htau'}$
  and $\tconsistent{\htau}{\htau'}$
  and $\htau\neq\htau'$.
\end{lem}

In all other cases, casts either consist of boxed values or
indeterminate terms according to the remaining rules
in \Figref{fig:isFinal}.
%
In particular, Rule \rulename{ICastHoleGround} handles casts that are not of
the form $\dcastthree{\dexp}{\htau_1}{\tehole}{\htau_2}$, from hole to
ground type.
%
In the cast calculus, since the only canonical form of type $\tehole$
is $\dcasttwo{\dexp}{\htau_1}{\tehole}$, there exist no such
irreducible terms.

\subsubsection{Type Safety} 
%
Type safety establishes that the static and dynamic semantics of a
language cohere.
%
We follow the approach developed by \citet{wright94:_type_soundness},
now standard \cite{pfpl}, which distinguishes two type safety
properties, preservation and progress.
%
To permit the evaluation of incomplete programs, we establish these
properties for terms typed under arbitrary hole context $\Delta$.
%
We assume an empty context for~$\hGamma$; to run open programs, the
system may treat free variables as empty holes with a corresponding
name.

@@

The progress theorem establishes that the dynamic semantics accounts
for every well-typed term, i.e. that we have not forgotten some
necessary rules or premises.
%
\begin{thm}[Progress]
  If $\hasType{\Delta}{\emptyset}{\dexp}{\htau}$ then either
  (a) $\stepsToD{}{\dexp}{\dexp'}$ or
  (b) $\isBoxedValue{\dexp}$ or 
  (c) $\isIndet{\dexp}$.
\end{thm}
\noindent
The key to establishing the progress theorem under a non-empty hole
context consists of explicitly accounting for indeterminate forms,
i.e. those rooted at either a hole closure or a failed cast.
%
The proof relies on canonical forms lemmas stated in
Appendix~\ref{sec:additional-defns}.

\subsubsection{Complete Programs} 
%
Although this paper focuses on running \emph{incomplete} programs, it helps
to know that its machinery does not interfere
with reasoning about the behavior of \emph{complete} programs, i.e. those
with no holes.
%
Appendix~\ref{sec:additional-defns} defines complete programs via judgements~$\isComplete{e}$ and~$\isComplete{d}$. 
%
Failed casts cannot appear in complete internal expressions. 
%
The following theorem establishes that expansion preserves program completeness.
\begin{thm}[Complete Expansion] ~
  \begin{enumerate}[nolistsep]
    \item
      If $\isComplete{\hexp}$
      and $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$
      then $\isComplete{\dexp}$.
    \item
      If $\isComplete{\hexp}$
      and $\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$
      then $\isComplete{\dexp}$.
  \end{enumerate}
\end{thm}

The following preservation theorem establishes that stepping preserves program completeness.\todo{can we do complete preservation without the typing assumption?}
\begin{thm}[Complete Preservation]
  If $\hasType{\hDelta}{\emptyset}{\dexp}{\htau}$
  and $\isComplete{\dexp}$
  and $\stepsToD{}{\dexp}{\dexp'}$
  then $\hasType{\hDelta}{\emptyset}{\dexp'}{\htau}$
  and $\isComplete{\dexp'}$.
\end{thm}

@@

The supplemental material includes our Agda-based
mechanization  \cite{norell2009dependently,norell:thesis} 
of the semantics and metatheory of \HazelnutLive, 
%including proofs of all of the 
including the theorems stated above, and all necessary lemmas\todo{make this so ian!}{}. 
%
%We choose the 
%(as did the mechanization of \Hazelnut by \citet{popl-paper}, though only a few definitions are common). 
%
%Agda is a good choice because it is designed to explicitly communicate a proof's structure, as is our goal, rather than relying on proof automation. 
%
%Agda itself was also an inspiration for this work because it supports holes, albeit in a more limited form than described here (cf. Sec.~\ref{sec:intro}). 
%
%
Our approach is standard: We model judgements as 
inductive datatypes, and rules as dependently typed constructors of these judgements\todo{cite something?}{}. 
%
We adopt Barendregt's convention for bound variables \cite{urban,barendregt84:_lambda_calculus} and encode typing 
contexts, hole contexts and hole environments using metafunctions.\todo{say something about axioms?}{}

>>>>

\parahead{Overview} The syntax of the core calculus given in Fig.~\ref{fig:hazelnut-live-syntax} consists of types and expressions with holes. 
We distinguish between {external} expressions, $e$, and {internal} expressions, $d$. 
External expressions correspond to programs as entered by the programmer 
(see Sec.~\ref{sec:intro} for discussion of implicit, manual, semi-automated and fully automated hole entry methods). 
Each well-typed external expression (as specified in Sec.~\ref{sec:external-statics} below) expands to a well-typed internal expression (see Sec.~\ref{sec:expansion}) before it is evaluated (see Sec.~\ref{sec:evaluation}). 
We take this approach, notably also taken in the ``redefinition'' of Standard ML by \citet{Harper00atype-theoretic}, because (1) the external language supports type inference and explicit type ascriptions, $\hexp : \htau$, but it is formally simpler to eliminate ascriptions and specify a type assignment system when defining the dynamic semantics; and 
(2) we need additional syntactic machinery during evaluation for tracking hole closures and dynamic type casts. 
This machinery is inserted by the expansion step, rather than entered explicitly by the programmer. 
In this regard, the internal language is analogous to the cast calculus in the gradually typed lambda calculus \cite{DBLP:conf/snapl/SiekVCB15,Siek06a}, though as we will see the \HazelnutLive internal language goes beyond the cast calculus in several respects. We have mechanized these formal developments using the Agda proof assistant \cite{norell:thesis,norell2009dependently} 
(see Sec.~\ref{sec:agda-mechanization}). Rule names, e.g. \rulename{SVar}, correspond to the names used in the mechanization. We have also implemented \HazelnutLive in a manner that maintains a novel end-to-end well-definedness invariant (see Sec.~\ref{sec:implementation}).

@@

Before describing the rules in detail, let us state two other guiding theorems. 
The following theorem establishes that an expansion exists for every well-typed external expression. 
The mechanization also establishes that ywhen an expansion exists, it is unique (not shown).
 \begin{thm}[Expandability] \label{thm:expandability}~
  \begin{enumerate}[nolistsep]
    \item
      If $\hsyn{\hGamma}{\hexp}{\htau}$
      then $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$
      for some $\dexp$ and $\Delta$.
    \item
      If $\hana{\hGamma}{\hexp}{\htau}$
      then $\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$
      for some $\dexp$ and $\htau'$ and $\Delta$.
  \end{enumerate}
\end{thm}
% \noindent
% The following theorem establishes that when an expansion exists, it is unique.
% \begin{thm}[Expansion Unicity] \label{thm:expansion-unicity}~
%   % \begin{enumerate}[nolistsep]
%   %   \item
%       If $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$
%       and $\expandSyn{\hGamma}{\hexp}{\htau'}{\dexp'}{\Delta'}$
%       then $\htau=\htau'$ and $\dexp=\dexp'$ and $\Delta=\Delta'$.
%   %   \item
%   %     If $\expandAna{\hGamma}{\hexp}{\htau_1}{\dexp}{\htau_2}{\Delta}$
%   %     and $\expandAna{\hGamma}{\hexp}{\htau_1}{\dexp'}{\htau_2'}{\Delta'}$
%   %     then $\dexp=\dexp'$ and $\htau_2=\htau_2'$ and $\Delta=\Delta'$.
%   % \end{enumerate}
% \end{thm}
% \noindent
The following theorem establishes that expansion generalizes external typing.
\begin{thm}[Expansion Generality] \label{thm:expansion-generality}~
  \begin{enumerate}[nolistsep]
    \item
      If $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$
      then $\hsyn{\hGamma}{\hexp}{\htau}$.
    \item
      If $\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$
      then $\hana{\hGamma}{\hexp}{\htau}$.
  \end{enumerate}
\end{thm}

@@

To recap, the result of expansion is a well-typed internal expression with appropriately initialized hole closures and casts. 
This section specifies the dynamic semantics of \HazelnutLive as a ``small-step'' transition system over internal expressions equipped with a meaningful notion of type safety even for incomplete programs, i.e. expressions typed under a non-empty hole context, $\Delta$. 
We establish that evaluation does not stop immediately when it encounters a hole, nor when a cast fails, by precisely characterizing when evaluation \emph{does} stop. 
We also establish that an essentially standard notion of type safety holds when running complete programs.

@@

Figures~\ref{fig:isGround}-\ref{fig:step} specify the dynamic semantics. 
%
Most of the cast-related machinery closely follows the cast calculus from
the ``refined'' account of the gradually typed lambda calculus
by \citet{DBLP:conf/snapl/SiekVCB15}, which is known to be
theoretically well-behaved. 
%
In particular, \Figref{fig:isGround} defines the judgement
$\isGround{\htau}$, which distinguishes the base type~$b$ and the
least specific arrow type~$\tarr{\tehole}{\tehole}$ as \emph{ground
types}; this judgement helps simplify the treatment of function casts, discussed below.

%
\Figref{fig:isFinal} defines the judgement $\isFinal{d}$, which
distinguishes the final, i.e. irreducible, forms of the transition system.
%
The two rules distinguish two classes of final forms: (possibly-)boxed values and
indeterminate forms.%
%
\footnote{
        In most accounts of the cast calculus, values and ground types
        are distinguished with separate grammars together with an
        implicit identification convention.
        %
        Our judgemental formulation is more faithful to the mechanization and
        cleaner for our purposes, because we are distinguishing several
        classes of final forms.
}
The judgement $\isBoxedValue{d}$ defines (possibly-)boxed values as
either ordinary values~(Rule~\rulename{BVVal}), or one of two cast forms: casts
between disequal function types and casts from a ground
type to the hole type. In each case, the cast must appear inductively on a boxed
value.
These forms are irreducible because they represent values that have been boxed 
but have never flowed into a corresponding ``unboxing'' operation, discussed below. 
%
%, which correspond to the values from the cast calculus and include
%the classic values from the lambda calculus, distingui%shed by
%$\isValue{d}$,
%
The judgement $\isIndet{d}$ defines \emph{indeterminate} forms, so
named because they are rooted at expression holes and
failed casts, and so, conceptually, their ultimate value
awaits programmer action (see Sec.~\ref{sec:resumption}).
%
The first two rules specify that {empty} hole closures are always
indeterminate, and that {non}-empty hole closures are indeterminate when
they consist of a {final} inner expression.
%
Below, we describe failed casts and the remaining indeterminate forms simultaneously
with the corresponding transition rules.\todo{relationship to weak head normal forms}

Figures~\ref{fig:instruction-transitions}-\ref{fig:step} define\todo{use define instead of specify everywhere?}{} the transition rules. 
%
Top-level transitions are \emph{steps}~$\stepsToD{}{d}{d'}$, governed by Rule~\rulename{Step} in \Figref{fig:step}, which
%
(1) decomposes $d$ into an evaluation context, $\evalctx$, and a selected sub-term, $d_0$;
%
(2) takes an \emph{instruction transition}, $\reducesE{}{d_0}{d_0'}$, as specified in \Figref{fig:instruction-transitions};
%
and (3) places $d_0'$ back at the selected position, indicated in
the evaluation context by the \emph{mark}, $\evalhole$, to obtain $d'$.%
\footnote{
        We say ``mark'', rather than the more conventional ``hole'', to avoid confusion with the (orthogonal) holes of \HazelnutLive.
        %
        %the form $\evalhole$ in the grammar of
        %evaluation contexts is referred to as the \emph{hole}, but
        %this hole is a technical device entirely orthogonal to the
        %holes of this paper, so we use the term ``mark'' instead.
        }
%
%% \matt{Next paragraph can be dropped entirely for space}
%% %
This approach was originally developed in the reduction semantics of \citet{DBLP:journals/tcs/FelleisenH92} and is the predominant style of operational semantics in the literature on gradual typing. 
Because we distinguish final forms judgementally, rather than syntactically, we use a judgemental formulation of this approach called a \emph{contextual dynamics} by \citet{pfpl}. 
It would be straightforward to construct an equivalent structural operational semantics \cite{DBLP:journals/jlp/Plotkin04a} by using search rules instead of evaluation contexts (\citet{pfpl} relates the two approaches).

The rules take care to maintain the property that final expressions cannot take a step.%
\todo{prove this as stated from Ian's disjointness proofs}{}%
\begin{thm}[Finality] There does not exist $d$ such that both $\isFinal{d}$ and $\stepsToD{}{d}{d'}$ for some $d'$.
\end{thm}

@@

\subsubsection{Application and Substitution} 
%
Rule \rulename{ITBeta} in Fig.~\ref{fig:instruction-transitions} specifies the
standard beta reduction transition.
%
The bracketed premises of the form $\maybePremise{\isFinal{\dexp}}$ in
Fig.~\ref{fig:instruction-transitions}-\ref{fig:step} may be \emph{included}
to specify an eager, left-to-right evaluation strategy, or \emph{excluded} to
leave the evaluation strategy and order unspecified.
%
In our metatheory, we exclude these premises, both for
the sake of generality, and to support the specification of the fill-and-resume operation~(see \Secref{sec:resumption}).

@@

Substitution, written $[d/x]d'$, operates in the standard capture-avoiding manner~\cite{pfpl} (see Appendix~\ref{sec:additional-defns}).
%
The only cases of special interest arise when substitution reaches a hole closure:
\[
\begin{array}{rcl}
  [d/x]\dehole{u}{\sigma}{} & = & \dehole{u}{[d/x]\sigma}{} \\%
  \substitute{d}{x}{\dhole{d'}{u}{\sigma}{}} & = & \dhole{[d/x]d'}{u}{[d/x]\sigma}{}
\end{array}
\]
In both cases, we write~$[d/x]\sigma$ to perform substitution on each expression in the hole environment~$\sigma$, i.e. the environment ``records'' the substitution. 
%
For example, $\stepsToD{}
    {\hap{(\halam{x}{b}{\halam{y}{b}{\dehole{u}{[x/x, y/y]}{}}})}{c}}
    {\halam{y}{b}{\dehole{u}{[c/x, y/y]}{}}}$. 
%
Beta reduction can duplicate hole closures.
%
Consequently, the environments of different closures with the same hole name may differ, 
e.g., when an reduction applies a function with a hole closure body multiple times as in Fig.~\ref{fig:grades-example}.%
\todo{show example here or refer back to sec 2?}{} 
%
Hole closures may also appear within the environments of other hole
closures, giving rise to the closure paths described in
Sec.~\ref{sec:paths}\todo{where?}{}.

The ITBeta rule is not the only rule we need to handle function
application, because lambdas are not the only final form of arrow type.
%
Two other situations may also arise.

@@

Second, the expression in function position may be indeterminate,
where arrow cast conversion is not applicable,
e.g. $\hap{(\dehole{u}{\sigma}{})}{c}$.
%
In this case, the application is indeterminate (Rule~\rulename{IAp}
in \Figref{fig:isFinal}), and the application reduces no
further.\todo{mention WHNF here?}

@@

\subsubsection{Casts}
Rule \rulename{ITCastId} drops identity casts. The remaining instruction
transition rules assign meaning to non-identity casts. 
%
As discussed in Sec.~\ref{sec:cast-insertion}, the structure of a term
cast to hole type is statically obscure,
%
so we must wait for \emph{use} of the term at some other type, via a
cast away from hole type, to detect the type error dynamically.
%
Rules \rulename{ITCastSucceed} and \rulename{ITCastFail} handle this situation when the
two types involved are ground types (Fig.~\ref{fig:isGround}).
%
If the two ground types are equal, then the cast succeeds and the cast
may be dropped.
%
If they are not equal, then the cast fails and the failed cast form,
$\dcastfail{\dexp}{\htau_1}{\htau_2}$, arises.
%
Rule \rulename{TAFailedCast} specifies that a failed cast is well-typed exactly
when $d$ has ground type $\tau_1$ and $\tau_2$ is a ground type
disequal to $\tau_1$.
%
Rule \rulename{IFailedCast} specifies that a failed cast operates as an
indeterminate form (once $d$ is final), i.e. evaluation does not stop.\todo{talk about blame somewhere}{}

Rules \rulename{TAFailedCast} and \rulename{IFailedCast} each operate at ground type. 
%
The two remaining instruction transition rules, Rule~\rulename{ITGround}
and~\rulename{ITExpand}, insert intermediate casts from non-ground type to a
consistent ground type, and \emph{vice versa}.
%
These rules serve as technical devices, permitting us to restrict our
interest exclusively to ground types.
%
Here, the only non-ground types are the arrow types, so the grounding
judgement~$\groundmatch{\tau_1}{\htau_2}$ (\Figref{fig:groundmatch}),
produces the ground arrow type~$\tarr{\tehole}{\tehole}$.
%
More generally, the following invariant governs this judgement.
\begin{lem}[Grounding] 
  If $\groundmatch{\htau_1}{\htau_2}$
  then $\isGround{\htau_2}$
  and $\tconsistent{\htau_1}{\htau_2}$
  and $\htau_1\neq\htau_2$.
\end{lem}

In all other cases, casts evaluate either to boxed values or to 
indeterminate forms according to the remaining rules
in \Figref{fig:isFinal}.
%
In particular, Rule \rulename{ICastHoleGround} handles casts that are not of
the form $\dcastthree{\dexp}{\htau_1}{\tehole}{\htau_2}$, from hole to
ground type.
%
(In the cast calculus, there exist no such
irreducible terms; the only canonical form of type $\tehole$
is $\dcasttwo{\dexp}{\htau_1}{\tehole}$.)

\subsubsection{Type Safety} 
%
The purpose of establishing type safety is to ensure that the static and dynamic semantics of a
language cohere.
%
We follow the approach developed by \citet{wright94:_type_soundness},
now standard \cite{pfpl}, which distinguishes two type safety
properties, preservation and progress.
%
To permit the evaluation of incomplete programs, we establish these
properties for terms typed under arbitrary hole context $\Delta$.
%
We assume an empty typing context, $\hGamma$; to run open programs, the
system may treat free variables as empty holes with a corresponding
name.

@@

The progress theorem establishes that the dynamic semantics accounts
for every well-typed term, i.e. that we have not forgotten some
necessary rules or premises.
%
\begin{thm}[Progress]
  If $\hasType{\Delta}{\emptyset}{\dexp}{\htau}$ then either
  (a) $\stepsToD{}{\dexp}{\dexp'}$ or
  (b) $\isBoxedValue{\dexp}$ or 
  (c) $\isIndet{\dexp}$.
\end{thm}
\noindent
The key to establishing the progress theorem under a non-empty hole
context is to explicitly account for indeterminate forms,
i.e. those rooted at either a hole closure or a failed cast.
%
The proof relies on canonical forms lemmas stated in
Appendix~\ref{sec:additional-defns}.

\subsubsection{Complete Programs} 
%
Although this paper focuses on running \emph{incomplete} programs, it helps
to know that its machinery does not interfere
with reasoning about the behavior of \emph{complete} programs, i.e. those
with no holes.
%
Appendix~\ref{sec:additional-defns} defines the judgements~$\isComplete{e}$ and~$\isComplete{d}$. 
%
Failed casts cannot appear in complete internal expressions. 
%
The following theorem establishes that expansion preserves program completeness.
\begin{thm}[Complete Expansion] ~
  \begin{enumerate}[nolistsep]
    \item
      If $\isComplete{\hexp}$
      and $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$
      then $\isComplete{\dexp}$.
    \item
      If $\isComplete{\hexp}$
      and $\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$
      then $\isComplete{\dexp}$.
  \end{enumerate}
\end{thm}

The following preservation theorem establishes that stepping preserves program completeness.
\begin{thm}[Complete Preservation]
  If $\hasType{\hDelta}{\emptyset}{\dexp}{\htau}$
  and $\isComplete{\dexp}$
  and $\stepsToD{}{\dexp}{\dexp'}$
  then $\hasType{\hDelta}{\emptyset}{\dexp'}{\htau}$
  and $\isComplete{\dexp'}$.
\end{thm}

@@

The supplemental material includes our Agda-based
mechanization  \cite{norell2009dependently,norell:thesis} 
of the semantics and metatheory of \HazelnutLive, 
%including proofs of all of the 
including the theorems stated above and necessary lemmas.
We take as assumptions only a few standard properties of general hypothetical judgements and finite maps.\todo{what about hole disjointness}
%
%We choose the 
%(as did the mechanization of \Hazelnut by \citet{popl-paper}, though only a few definitions are common). 
%
%Agda is a good choice because it is designed to explicitly communicate a proof's structure, as is our goal, rather than relying on proof automation. 
%
%Agda itself was also an inspiration for this work because it supports holes, albeit in a more limited form than described here (cf. Sec.~\ref{sec:intro}). 
%
%
Our approach is standard: we model judgements as 
inductive datatypes, and rules as dependently typed constructors of these judgements\todo{cite something?}{}. 
%
We adopt Barendregt's convention for bound variables \cite{urban,barendregt84:_lambda_calculus} and encode typing 
contexts, hole contexts and hole environments using metafunctions.\todo{say something about funext?}{}
